% Declarations for Front Matter

\title{Finding Critical and Gradient-Flat Points of Deep Neural Network Loss Functions}
\author{Charles Frye}
\degreesemester{Spring}
\degreeyear{2020}
\degree{Doctor of Philosophy}
\chair{Professor Michael R DeWeese}
\othermembers{Adjunct Assistant Professor Kristofer E Bouchard \\
	Professor Bruno A Olshausen \\
	Assistant Professor Moritz Hardt}
\numberofmembers{4}
\field{Neuroscience}
% This is optional
\campus{Berkeley}

\maketitle
% Delete (or comment out) the \approvalpage line for the final version.
\approvalpage{}
\copyrightpage{}

\include{other/frontmatter/abstract}

\begin{frontmatter}

\begin{dedication}
\null\vfil
\begin{center}
	To the victims of empires, hegemonies, and masters,
	large and small, visible and invisible,
	across time and around the world.
	A better tomorrow is possible.\\\vspace{12pt}
\end{center}
\setlength{\epigraphwidth}{0.4\linewidth}
	\epigraph{%
	I found it hard, it's hard to find.\\
	Oh well, whatever, nevermind.}%
	{Kurt Cobain,\\\emph{Smells Like Teen Spirit}, May 1991}
\vfil\null\
\end{dedication}

% You can delete the \clearpage lines if you don't want these to start on
% separate pages.

\tableofcontents
\clearpage
\listoffigures
\clearpage
\listofalgorithms{}
\addcontentsline{toc}{chapter}{List of Algorithms}

\begin{acknowledgements}
	The first and greatest debt I owe is to my mother,
	Dr.\ Anne Teresa Gearhart.
	She nurtured my love of inquiry, knowledge, and education,
	fighting to preserve it in a harsh and unwelcoming
	environment.
	This degree is the culmination of that love
	and so the product of hers.
	Against her heart but not her judgment,
	she let me leave home at just 15
	to attend the boarding school of my dreams,
	the Illinois Math and Science Academy,
	and obtain the space to grow,
	at times away, but never too far.
	Mom: I love you and appreciate everything you have done for me.
	I will do my best to return the favor and to make you proud.

	She also raised my two wonderful siblings, Teddy and Willow.
	Without them I would be a much less full person.
	Willow: here's to at least three more decades
	of disputation and cardiovascular exercise.
	Teddy: I miss you every day,
	even if we disagree about \emph{Star Wars: The Last Jedi}.

	As an undergraduate at the University of Chicago,
	I was blessed with incredible scientific mentorship,
	the value of which only becomes clearer with time.
	Dr.\ Margaret Wardle oversaw my undergraduate
	honors thesis and showed me how to be both
	a scientist and a full human being.
	Megin, I owe you my appreciation
	for the importance of a rigorous statistical method.
	Thanks for never asking why I was so interested
	in studying the psychopharmacology of party drugs.
	I still tell people about CyberBall.
	I worked simultaneously with
	Dr.\ Jason MacLean on a project that became
	the first scientific venture over which I felt
	full intellectual ownership.
	Jason, thanks for taking a chance on an
	an undergraduate with more enthusiasm than sense
	who maybe shouldn't have been given control of a laser.
	You are my model of integrity,
	as a scientist and a person,
	in the face of immense pressure to publish
	flashy, shoddy results.

	When I was deciding where to go to graduate school,
	Jason told me to ask myself:
	\enquote{where do the students seem most happy?
	Where is there a community I can see myself thriving in?}.
	I chose to do my PhD at Berkeley largely on that basis,
	and I can look back and see that the choice was the right one.
	I have been challenged to grow as a person
	and guided in that growth by the amazing community
	of students at Berkeley.

	First, Dylan \enquote{ChicknMcGizzard} Paiton, PhD.
	From my first rotation to our latest paper,
	we have been a productive research team.
	From riding the bus at the Ward House
	to building yurts at the Dog House,
	we have been partners in crime.
	To more science and more crimes!

	This thesis would not have been possible
	without Neha Wadia,
	who first introduced me to the problem
	of finding critical points of neural networks
	and who co-authored the papers
	that form the third chapter of this thesis.
	Neha, your wisdom, intellect, and integrity
	are an inspiration.

	From my first days at UChicago
	to these last days at Berkeley,
	I have counted on the friendship and scholarship
	of Ryan Zarcone.
	Ryan, thank you for always being down
	to listen to a half-finished rant
	about mathematics,
	to give presentations, blog posts,
	and manuscripts trenchant feedback,
	and to grab some Taco Bell or a whiskey
	and chew the fat.
	To a life of Epicurean delights
	and Platonic contemplation
	in equal measure!

	Dr.s Alexander Naka and Cameron Baker
	have been the Bubbles and Blossom
	to my Buttercup
	when they weren't busy being
	the SpongeBob and Patrick
	to my Squidward.
	Alex, even if you never convince me
	to that neuroscience is not butts,
	you will forever be one of my closest friends.
	Cam: thank you for keeping me ground,
	running buddy.

	If not for Dr.\ Nicholas Ryder asking me,
	casually over a beer,
	how I might go about computing a square root,
	the second chapter of this thesis would be much impoverished.
	Nick, I treasure our conversations
	at least as much as I do the free food at OpenAI\@.

	Among a blessed surplus of fabulous graduate student
	role models ahead of me in the program,
	including Dr.\ Brian Isett and Dr.\ Timothy Day,
	two stand out for the special contributions they made to
	my education in particular.
	Dr.\ Jesse Livezey,
	you have for years been the first person I go to
	when a quandary of ethics presents itself.
	Your and Sarah's commitment
	to justice and to compassion,
	even when it means hard work,
	is a bright light in a world with too much darkness.
	Dr.\ Natalia Bilenko,
	your light burns just as bright,
	and you were the best manager I ever had.
	Good luck and godspeed as you bring
	the machine learning community
	kicking and screaming into
	accepting its responsibility.

	It was an absolute privilege to
	be a part of the community of the
	Redwood Center for Theoretical Neuroscience.
	The special nature of Redwood as an
	interdisciplinary home for heretical
	neuroscientists was the other major factor
	that led me to choose to do my thesis work at Berkeley.
	My committee members,
	Dr.s Kristofer Bouchard,
	Michael DeWeese,
	and Bruno Olshausen,
	imbue the community with a peculiar sense of purpose:
	to question everything and demand nothing short
	of the right answer.
	It was in that heterodox spirit that
	the work in this thesis was conducted.
	Dr.\ Yubei Chen's and Brian Cheung's commitments
	to scholarship, both fundamental and cutting edge,
	were direct role models.
	Pratik Sachdeva and Vasha DuTell,
	I'll catch you at the next stretch break.
	Michael Fang: if you need me,
	I'll be in my rabbit hole.
	Charles, thanks for the desk.

	The community at Redwood helped me
	to grow into the scholar I am today.
	But that growth would not have been possible
	without the freedom afforded to me by the
	National Science Foundation's
	Graduate Research Fellowship,
	which funded me through half of my PhD,
	and the Berkeley Chancellor's Fellowship,
	which supported me in the first two years.
	I am deeply appreciative
	of the public's investment in me
	and in basic science,
	and I hope to repay it as best I can.

	Looking outside the narrow, at times myopic,
	world of research,
	I would like to thank all of my wonderful friends,
	whose lives and passions and stories
	deeply enrich my own.
	To the saviors of Isger,
	the heroes who closed the Worldwound,
	and the adventurers who defeated the Forgotten Pharaoh Cult,
	I look forward to rolling strange polyhedra
	with you all a whole lot more in the future.
	To the members of the Hot Pony Social Club,
	both past and future,
	Daniel Mossing, Sara Popham, Storm Slivkoff,
	Sophie Obayashi, Dr.\ Nicholas Jourjine,
	and Dr.\ Kevin Doxzen:
	living with you was a pleasure and a privilege.
	To the VeryBestFriends,
	especially those who haven't been mentioned yet,
	Kelsey Howard, William Shainin, Christine Hopkins,
	Paul Cullen, Andres Vargas, Neil Wilson, Katherine Cording,
	Parker Conroy, and Maxine Marshall:
	I cherish you.

	I want also to thank my partners, past and present,
	especially Justyn, Cate, and Venatius.
	Though I did not and do not always show it
	as clearly as I'd like,
	I care deeply for you.
	Thank you for tolerating my absorption in my work
	as much as you did.


\end{acknowledgements}

\end{frontmatter}
